Unsupervised training of maximum-entropy models for lexical selection in rule-based machine translation

### 2015-02-09 2015-eamt-melsa/2015-eamt-melsa.tex
glue in melsa paper
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@58776 72bbbca6-d526-0410-a7d9-f06f51895060

%
% File eamt15.tex, an adaptation of eamt14.tex (which was a copy of
% eamt12.tex)
%
% Contact: eamt2015@dlsi.ua.es

%%% To ease future customizations, various replaceables have been paramaterized
%%% as listed in the newcommands section

@@

@@

\title{Instructions for \confname\ Proceedings}

\author{First Author\\
  Affiliation / Address line 1\\
  Affiliation / Address line 2\\
  Affiliation / Address line 3\\
  {\tt email@domain}  \And
  Second Author\\
  Affiliation / Address line 1\\
  Affiliation / Address line 2\\
  Affiliation / Address line 3\\
  {\tt email@domain}}

@@

\begin{document}
\maketitle
\begin{abstract}
   ... 
\end{abstract}

@@

\cite{koehn2005}

\bibliographystyle{apalike}

\bibliography{2015-eamt-melsa}

>>>>

% test

@@

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{multirow}
\usepackage{url}
\usepackage{natbib}
\usepackage{color}
\usepackage[small,bf]{caption}

\usepackage[backgroundcolor=lightgray]{todonotes}
%\usepackage[disable]{todonotes}

@@

% \newcommand{\comment}[1]{{\small{\texttt{$\clubsuit$#1$\clubsuit$}}}}
%\newcommand{\comment}[1]{\marginpar{\scriptsize\sf \textcolor{blue}{#1}}}
\newcommand{\comment}[1]{\todo{#1}}

  \title{Unsupervised training of maximum-entropy models for lexical
    selection in rule-based machine translation}

%\author{}

\author{
 Francis M.\ Tyers \\
 HSL-fakultehta, \\
 UiT Norgga \'{a}rktala\v{s} universitehta, \\
 N-9018 Romsa \\
 \and
 Felipe S\'anchez-Mart\'inez \\
 Dept.\ Lleng.\ i Sist.\ Inform., \\
 Universitat d'Alacant,\\
 E-03071 Alacant\\
 \and
 Mikel L.\ Forcada \\
 Dept.\ Lleng.\ i Sist.\ Inform., \\
 Universitat d'Alacant,\\
 E-03071 Alacant\\
}

@@

\begin{document}
\maketitle

\begin{abstract}
  This article presents a method of training maximum-entropy models to
  perform lexical selection in a rule-based machine translation
  system. The training method described is unsupervised; that is, it
  does not require any annotated corpus. The method uses
  source-language monolingual corpora, an existing machine translation
  (MT) system and a statistical target-language model. Using the
  existing MT system, the sentences in the source-language corpus are
  translated in all possible ways according to the different
  translation equivalents in the bilingual dictionary of the system.
%  taking into account all of the lexical selection possibilities.
  These translations are then scored on the target-language model and
  the scores are normalised to provide fractional counts for training
  a source-language maximum-entropy lexical-selection model. We show
  that this model can perform equally well, or better, than using the
  target-language model directly for lexical selection, at a
  substantially reduced computational cost.
\end{abstract}

@@

Corpus-based MT has been the primary research direction in the field 
of machine translation in recent years. However, RBMT systems are 
still being developed, and there are many successful commercial 
and non-commercial systems. One reason for the continued development
of RBMT systems is that in order to be successful, corpus-based MT requires parallel corpora in
the order of tens of millions of words. Although for some language
pairs these exist, they only exist for a fraction of the 
world's languages.

An RBMT system typically consists of an analysis component,\footnote{For example
a morphological or syntactic analyser.} a transfer
component and a generation component. As part of the transfer component
it is necessary to make choices regarding words in the source language
which may have more than one translation in the target language.

Lexical selection is the task of choosing, for a given source-language
word, the most adequate translation in the target language among a
known set of alternatives.  The task is related to the task of
word-sense disambiguation \citep{ide98a}. However, it is different to
word-sense disambiguation in that lexical selection is a bilingual
problem, not a monolingual problem: its aim is to find the most
adequate translation, not the most adequate sense. Thus, it is not
necessary to choose between a series of fine-grained senses if all
these senses result in the same final translation; however, it may be
necessary to choose a different translation for the same sense, for
example in a collocation.
% It could also be the case that a single sense could have more than
% one possible translation (for example synonyms).

%\comment{FSM: A more general introduction to MT and RBMT should be
%  provided. The necessity of the method should also be justified (lack
%  of parallel corpora), probably at the same time RBMT is claimed to
%  be a need for most language pairs with no parallel corpora
%  available.}

\subsection{Prior work}
\comment{This section should be ironed out a bit.}  In their
pioneering work, \cite{dagan1994word} used the term
\emph{word sense disambiguation} to refer to what is actually target
word selection in machine translation; they used a parser to identify
some syntactic relations such as subject--object or subject--verb.
After generating all the possible translations for a given
input sentence using an ambiguous bilingual dictionary, they extract the syntactic
tuples from the target language and count the frequency in a previously-trained target-language 
model of tuples.  They use maximum-likelihood estimation to calculate the probability of a given 
target language tuple being the translation of a given source language tuple, with
an automatically determined threshold for confidence.

Then \cite{berger1996} illustrated the use of maximum-entropy classifiers
on the specific problem of lexical selection in IBM-style word-based
statistical machine translation. Other authors
\citep{carbonell06,melero07a} have used target-language models to rank
the translations resulting from all possible combinations of lexical
selections.  Nowadays, in state-of-the-art phrase-based statistical
machine translation \citep{koehn2009statistical}, lexical selection is
taken care of by a combination of the translation model and the
language model. The translation model provides probabilities of
translation between words or word sequences (often referred to as
\emph{phrases}) in the source and target language. The target-language
model provides probabilities of word sequences in the target language.
As regards rule-based machine translation, \cite{sanchez07a} proposed 
to use information from the target language to train a statistical 
lexical-selection module but did not provide any experiments. 
\todo[inline]{FMT: integrate \citep{sanchez07a} somewhere here}
\cite{marechek10} trained a maximum-entropy lexical selector for their TectoMT rule-based system
using a bilingual corpus.  More recently, \cite{tyers12a} presented a
method of lexical selection for rule-based machine translation based
on rules which select or remove translations in fixed-length contexts,
along with a training method for learning the rules from a
word-aligned parallel corpus. The principal disadvantage with this
method is that it relies on a parallel corpus for the language pair,
something which is not always available.

The remainder of the article is organised as follows: 
\todo[inline]{FMT: Add article structure}

\section{Method}

%\todo[inline]{FSM: In the experiments a development corpus is
%  mentioned. For what is it used? This should be explain in this
%  section.}

\comment{Improve this} Lexical selection in this paper considers for
each word a simple source-language context made up of neighbouring
lemma+part-of-speech combinations. Contexts considered include up to
two words to the left and up to two words to the right of the word to
be translated. 
%\comment{FSM: Later, a development corpus is
%  mentioned. In this section it should be mentioned what it is used
% for.}

Let the probability of a translation $t$ being the translation of a
word $s$ in an SL context $c$ be $p_s(t|c)$. In principle, this value
could be estimated directly from the available corpora for every
combination of $(s, t, c)$.  This would however present two questions:
(1) how should the relevant contexts be chosen? and (2) what should be
done when the translations of words are not found in the corpus?  A
maximum-entropy model answers both of these questions. It allows the
contexts that we consider to be linguistically interesting to be
defined \emph{a priori} and then integrate these seamlessly into a
probabilistic model \citep{Manning99b}.  In answer to the second
question, a maximum-entropy model assumes nothing about what is not
present in the training data. That is, if there is no information in
the training data, then it assumes that all outcomes ---that is, all
possible translations--- are equally likely.  As previously mentioned,
the principle of maximum entropy has been applied to the problem of
lexical selection before; in particular, \cite{berger1996} cast the
problem of lexical selection in statistical MT as a classification
problem. They learn a separate maximum-entropy classifier for each SL
word form,\comment{[:we do too, integrate this later]} using SL context to
distinguish between possible translations. These classifiers are then
incorporated into the translation model of their word-based SMT
system.\comment{[how is our work different?]}  In their approach, a
classifier consists of a set of binary feature functions and
corresponding weights for each feature. 
In both \cite{berger1996} and our method, features are defined in the
form $h_k^{s}(t, c)$, where $t$
is a translation, and $c$ is an SL context. One difference is that \cite{berger1996} take $s$, $t$
and $c$ to be based on word forms, whereas in our method they are based
on lemma forms.
A feature where the Spanish word \emph{pez} (`fish' as a living animal) is seen as the
translation of \emph{arrain} (`fish') in the context \emph{arrain
  handi} `big fish' would therefore be defined as:
%\comment{FSM: I have introduced the index $k$
%  because several features are defined for a SL word $s$.} 
\begin{equation}
h_k^{\mathrm{arrain}}(t, c) = \left \{ \begin{matrix}
                 ~~1 & \mbox{if } t~ =~ \textit{pez}~ \mathrm{and}~ \textit{handi}~ \mathrm{follows}~ \textit{arrain}   \\
                 ~~0 & \mbox{otherwise} \end{matrix} \right.
\label{eq:feat-func}
\end{equation}
This feature considers a context of zero words to the left of the
problem word and one word to the right of it.

As a result of training, each feature in the classifier is assigned a
weight $\lambda_k^s$, and combining these weights of active features
as in equation~(\ref{eq:max-ent-sum}) yields the probability of a
translation $t$ for word $s$ in context $c$.

\begin{equation}
p_s(t|c) = \frac{1}{Z^s(c)} \exp{ \sum_{k=1}^{n_F} \lambda_k^s h_k^s(t, c)}
\label{eq:max-ent-sum}
\end{equation}

In this equation, $Z^s(c)$ is a normalising constant. Thus, the most
probable translation \(t^\star\) can be found using
% equation~(\ref{eq:max-ent-max}).

\begin{equation}
  t^\star=\underset{t \in T_s(s)}{\arg\max} ~ p_s(t|c) = \underset{t \in T_s(s)}{\arg\max} \sum_{k=1}^{n_F} \lambda_k^s h_k^s(t, c) .
\label{eq:max-ent-max}
\end{equation}

The approaches by \cite{berger1996} and by \cite{marechek10} cited
above both take advantage of a parallel corpus to collect counts of
contexts and translations in order to train the maximum-entropy
model. However, parallel corpora are not available for the majority of
the world's written languages. In this section we describe an
unsupervised method to learn the model using only monolingual corpora
and the components from an existing machine translation system.

The input to our method consists of a collection of samples,
$\mathcal{G} = (S, G)$, where $S = (s_1, s_2, \ldots s_{|S|})$ is a
sequence of source-language words, and $G = \{g_1, g_2, \ldots
g_{|G|}\}$ is a set of possible \emph{lexical-selection paths}. A
lexical-selection path $g = (t_1, t_2, \ldots, t_{|s|})$ is a sequence
of lexical-selection choices of those SL words, where, \(t_i=T(s_i)\)
and \(T(\cdot)\) is the set of all translations for a source
word. This is produced in the first stages of rule-based machine
translation, just after morphological analysis, part-of-speech
tagging, and bilingual dictionary lookup, and before any structural
transfer takes place (we will call this \emph{pre-lexsel}). In our
model, it is after these first stages that lexical selection
(\emph{lexsel}) occurs. After lexical selection, structural transfer
and generation take place; a function $\tau(g_i,S)$ represents the
result of these last stages, which we will call \emph{post-lexsel},
and returns a finished translation of a specific lexical-selection
path $g_i$ of sentence $S$. Schematically:
\begin{displaymath}
  \label{eq:steps}
  S \to \mbox{\framebox{\parbox{0.8cm}{\textit{pre- lexsel}}}} \to (\{g_i\}_{i=1}^{i=|G|},S) \to \mbox{\framebox{\parbox{1.0cm}{\textit{\textbf{lexsel}}}}} \to (g^\star,S) \to \mbox{\framebox{\parbox{0.8cm}{\textit{post- lexsel}}}} \to \tau(g^\star,S)
\end{displaymath}

%Taking our previous example for the parallel corpus, Table~\ref{table:ex-monocorp} 
%presents an example of five possible samples.

As our method is unsupervised, and therefore the occurrences of
specific lexical selection events \((s,t,c)\) cannot be counted, a
target language model \(P_{TL}(\cdot)\) is used to compute a value for
the fractional count for disambiguation path \(g_i\), $p(g_i|S)$ after
suitable normalisation:
\begin{equation}
 \label{eq:normalising}
   p(g_i|s) = \frac{P_{TL}(\tau(g_i,S))}{\sum_{g_i \in G} P_{TL}(\tau(g_i,S))}
\end{equation}
~\\
The maximum-entropy model is trained instead using the fractional count
$p(g_i|S)$ for the events $(s,t,c)$ found in \(g_i\), that is, when in
\(g_i\) $t$ is the translation for \(s\) in context \(c\). That is, as
if event $(s,t,c)$ had been seen a fractional number $p(g_i,s)$ of
times. We prune features occurring less than a certain number of times 
in the corpus, using a development corpus to guide pruning (see section~\ref{sec:eval-ref-results}).
\comment{Reword this to avoid repetition} The method used here
for lexical selection is analogous to the method used by
\cite{sanchez08b} to train a hidden-Markov-model-based part-of-speech
tagger in a rule-based machine translation system. 
Table~\ref{tb:workedexample} shows how a collection of samples
\(\mathcal{G}\) for the translation of Basque sentences containing the
word \emph{arrain} into Spanish.\footnote{The word \emph{arrain} `fish' has two translations
in Spanish, the first \emph{pez} refers to the living animal, whereas
\emph{pescado} refers to the animal as food.} 
%\comment{FSM: The correct translations
%  should be clearly identified in the table.}
%\todo[inline]{FSM: There is an important difference between the
%  approach in \cite{sanchez08b} and the one in this paper: we do not
%  have to face the problem of free rides. Should this be mentioned
%  here?}

\begin{table}
 \begin{center}
  \begin{tabular}{|c|l|r|}
    \hline
       & \textbf{Sentence} & $p(g_i|S)$ \\
    \hline % Los pescados grandes peque\~{n}o come
     \(S_1\) & \emph{Arrain handiak txikia jaten du}  &  \\
                        $\tau(g_{11},S_1)$ & El pez grande se come el peque\~{n}o &  0.830 \\
                        $\tau(g_{12},S_1)$ & El pescado grande se come el peque\~{n}o & 0.134 \\
                        $\tau(g_{13},S_1)$ & El pescado capaz se come el peque\~{n}o & 0.034\\
                        $\tau(g_{14},S_1)$ & El pez capaz se come el peque\~{n}o  & 0.001\\
    \hline % El padre el pescado nos ha preparado para cenar
     \(S_2\) & \emph{Aitak arraina prestatu digu afaltzeko} &  \\
                        $\tau(g_{21},S_2)$ & Padre nos ha preparado pescado para cenar & 0.846 \\
                        $\tau(g_{22},S_2)$ & Pap\'{a} nos ha preparado pescado para cenar & 0.134 \\
                        $\tau(g_{23},S_2)$ & Padre nos ha preparado pez para cenar & 0.015 \\
                        $\tau(g_{24},S_2)$ & Pap\'{a} nos ha preparado pez para cenar & 0.005 \\
    \hline % Aqu\'{\i} el pescado muy es dulce
    $S_3$  & \emph{Hemen arraina oso goxoa da}  & \\
                        $\tau(g_{31},S_3)$ & Aqu\'{\i} el pescado es muy rico & 0.912 \\
                        $\tau(g_{32},S_3)$ & Aqu\'{\i} el pescado es muy suave & 0.067 \\
                        $\tau(g_{33},S_3)$ & Aqu\'{\i} el pez es muy rico & 0.011 \\
                        $\tau(g_{34},S_3)$ & Aqu\'{\i} el pez es muy suave & 0.008 \\
                        $\tau(g_{35},S_3)$ & Aqu\'{\i} el pez es muy dulce & 0.001 \\
                        $\tau(g_{36},S_3)$ & Aqu\'{\i} el pescado es muy dulce & 0.001 \\
    \hline % Un pescado grande del mar es
    \(S_4\) & \emph{Itsasoko arrain handi bat da} & \\
                        $\tau(g_{41},S_4)$ & Es un pez grande del mar & 0.595 \\
                        $\tau(g_{42},S_4)$ & Es un pescado grande del mar & 0.403 \\
                        $\tau(g_{43},S_4)$ & Es un pez capaz del mar & 0.001 \\
                        $\tau(g_{44},S_4)$ & Es un pescado capaz del mar & 0.001 \\
    \hline % El pescado grande por esos serv\'{\i}an para alimentar
    $S_5$ &  \emph{Arrain handiak horiez baliatzen ziren elikatzeko} &  \\
                        $\tau(g_{51},S_5)$ & Los peces grandes se nutr\'{\i}an de esos & 0.862 \\
                        $\tau(g_{52},S_5)$ & Los pescados grandes se nutr\'{\i}an de esos & 0.121 \\
                        $\tau(g_{53},S_5)$ & Los peces grandes se alimentaban de esos & 0.012 \\
                        $\tau(g_{54},S_5)$ & Los peces capaces se alimentaban de esos & 0.001 \\
                        $\tau(g_{55},S_5)$ & Los pescados grandes se alimentaban de esos & 0.001 \\
                        $\tau(g_{56},S_5)$ & Los pescados capaces se alimentaban de esos & 0.001 \\
                        $\tau(g_{57},S_6)$ & Los peces capaces se nutr\'{\i}an de esos & 0.001 \\
                        $\tau(g_{58},S_7)$ & Los pescados capaces se nutr\'{\i}an de esos & 0.001 \\
    \hline

  \end{tabular}
  \caption{A sample of source-language sentences, the output of the
    possible lexical-selection paths after scoring on a
    target-language model. The scores are normalised as fractional
    counts, which are used as inputs to the maximum-entropy training
    process. In all sentences, the resultant translation with the highest share of the 
    probability mass is the most adequate.}
   \label{tb:workedexample}
 \end{center}
\end{table}

\section{Experimental setting}

This section describes the training and evaluation settings used in
the remainder of this paper. The primary motivation behind the
evaluation is that it should be automatic, meaningful, and be performed over a test set which
is large enough to be representative. It should evaluate both
performance on the specific subtask of lexical selection, and on the
whole translation task. Evaluating lexical-selection performance is an
intrinsic module-based evaluation.  It measures how well the lexical
selection module disambiguates the lexical-transfer
output as compared to a gold-standard corpus. The lexical 
transfer output is the result if looking up the translations of the 
source-language \emph{lexical forms} --- combinations of lemma and morphological
tags --- in the bilingual dictionary.

The
whole translation task evaluation is an extrinsic evaluation, which
tests how much the system improves as regards final translation
quality in a real system.

The lexical-selection module should be as language-independent as
possible. To that end, the language pairs tested should show as wide a
variety of linguistic phenomena. It is also important that the
methodology be as applicable to lesser-resourced and marginalised
languages, as to major languages.

This section begins with a short description of the Apertium 
platform \citep{forcada2011apertium}. This is followed by 
an overview of each of the language
pairs chosen for the evaluation. The corpora to be used for training
and evaluation will subsequently be described, along with the method
used for annotating them. This is followed by a description of the
performance measures to be used in the evaluation, and the reference
results using these metrics for each of the language pairs.

\subsection{Apertium}
\label{sec:apertium}

Apertium is a free/open-source rule-based machine translation platform, 
it comprises an engine, a toolbox and data to build rule-based machine
translation systems. Translation is implemented as a pipeline consisting
of the following modules: morphological analysis, morphological disambiguation,
lexical transfer, lexical selection, structural transfer and morphological
generation. The modules of the system for this method are the \emph{lexical transfer}
module which provides for each source language lexical form a set of possible
target language forms and the \emph{structural transfer} and \emph{morphological generation}
modules which for each pair of source language lexical form and translation 
produce a final translation in the target language.

\subsection{Language pairs}
\label{sec:eval-systems}

Evaluation will be performed using four language pairs. These pairs
have been selected as they include languages with different
morphological complexity, and different amounts of resources available
--- although for all pairs there is a parallel corpus available, which
is used for evaluation (see Section \ref{ss:measures}). Table~\ref{table:lang-pair-stats} 
presents statistics about the development status of each of the language 
pairs chosen, including the number of bilingual dictionary entries, coverage of
test corpora and translational ambiguity of the bilingual dictionary. Translation
ambiguity is provided for both all words, and for only ambiguous open-category 
words. In the experiments, lexical-selection rules are learnt for the open categories
of adjective, noun and verb.

\begin{table}
  \centering
  \begin{tabular}{|l|r|r|r|r|r|}
    \hline
    \multirow{2}{*}{\textbf{Pair}} & \textbf{Lexical} & \multirow{2}{*}{\textbf{Ambig.}} & \textbf{Na\"{i}ve}& \textbf{Total} & \textbf{Ambig. of}\\\
                          & \textbf{items} &                            & \textbf{coverage} & \textbf{ambig.} & \textbf{ambig. words}\\
    \hline
    \texttt{br-fr} & 27,690 & 235  &  94.47 \%  & 1.02 & 2.5 \\
    \texttt{mk-en} & 33,229 & 1,323 & 92.17 \%  & 1.37 & 2.8 \\
    \texttt{eu-es} & 19,323 & 455 & 91.70 \% & 1.08 & 2.4 \\
    \texttt{en-es} & 30,186 & 822 & 98.08 \% & 1.05 & 2.2 \\
    \hline
  \end{tabular}
  \caption{Statistics about development status of language pairs. The second
    column gives the number of entries in the bilingual dictionaries of the systems;
    the third column gives the total number of source language words with more 
    than one translation in the target language; the fourth column gives the coverage
    of the dictionaries on the test corpus (see section~\ref{sec:eval-corpus}); the fifth
    column gives the total mean ambiguity for open classes (nouns, verbs, adjectives);
    and the final column gives the mean ambiguity for ambiguous words in open classes.}
  \label{table:lang-pair-stats}
\end{table}

%\todo[inline]{FSM: More information about the language pairs needs to be
%  given: naive coverage, average lexical ambiguity over all words and
%  over the ambiguous ones (should this information be given for the
%  different lexical categories?). This information would allow the
%  reader to have an idea of how hard is the lexical selection problem
%  in each case.}

%\todo[inline]{FSM: If I remember correctly, we only deal with lexical
%  ambiguity in certain lexical categories. This is not mentioned
%  anywhere.}

\begin{description}
\item[Breton--French:] Development of the Breton--French pair has been
  described in two articles \citep{tyers09b,tyers10b}. The bilingual
  dictionaries were not built with polysemy in mind from the outset,
  but some entries were added later to start work on lexical
  selection. The version used in this paper is SVN revision
  41375.\footnote{\url{https://svn.code.sf.net/p/apertium/svn/trunk/apertium-br-fr}}
  This is the only available machine translation system between Breton
  and French. It has been developed part-time over a number of years.

\item[Macedonian--English:] The Macedonian--English pair
  in Apertium was created specifically for the purposes of running
  lexical-selection experiments. The resources reused from other pairs
  were the English morphological dictionary from the Icelandic and
  English pair \citep{brandt11}, the Macedonian morphological
  dictionary and constraint grammar from the Macedonian and Bulgarian
  pair \citep{rangelov11}, and the SETimes parallel corpus
  \citep{tyers10}.  The work was carried out over a period of eight
  days, and consisted of creating the bilingual dictionary and
  transfer rules. The bilingual dictionary was created by tagging both
  sides of the parallel corpus, word-aligning them with {\sc giza++}
  \citep{och03a} and extracting the probabilistic lexicon. Entries
  from this lexicon were checked manually according to frequency and
  included in the bilingual dictionary of the machine translation
  system. The most probable entry was checked to ensure that it was 
  an adequate translation, and if so marked as the 
  default.\footnote{A bilingual dictionary in Apertium
    \citep{forcada2011apertium} may contain more than one possible
    target-language translation for a given source-language
    word. Where there is more than one possible translation, the
    person who is writing the dictionary chooses the most general or
    most frequent translation among the set of possible translations
    and marks it as the \emph{linguistic default} translation.} As a
  result of attempting to include all possible translations, the
  average number of translations per word is much higher (see table~\ref{table:lang-pair-stats})
than in other pairs.  The transfer rules were written
  by hand to treat translation problems in the corpus. The version of
  the software used in this paper is SVN revision
  41476.\footnote{\url{https://svn.code.sf.net/p/apertium/svn/trunk/apertium-mk-en}}.
\item[Basque--Spanish:] The development of the Basque to Spanish pair
  is described in \cite{ginesti09}. The bilingual dictionary was taken
  from the free/open-source Matxin system \citep{mayor2011}. 
  Alternative translations were included in the bilingual
  dictionary.% , but non-default translations were marked with a
%  direction restriction\comment{FSM: This needs to be carefully
%    explained to avoid confusing the reader.}.\footnote{The bilingual
%    dictionaries in Apertium are bidirectional. They contain entries
%    which can be used for translating both from language $A$ to
%    language $B$ and vice versa. A direction restriction is used to
%    limit the entry to being used for translation in one translation
%    direction.} For lexical-selection experiments these direction
%  restrictions were removed.

\item[English--Spanish:] The English--Spanish pair was developed from
  a combination of the English--Catalan and Spanish--Catalan pairs
  over a period of around 3--5 months. The pair 
%has been used in other
%  lexical selection experiments \citep{sanchez07a}\comment{FSM: Sure?
%    It was on en-ca and in that paper no experimental results are
%    shown. This paper would be better mentioned in the introduction
%    (it mentions the use of TL info for lexical selection.},
  contains a number of entries in the bilingual dictionary with
  more than one translation. The most-frequent translation, as judged
  by the language pair developer, is marked as the \emph{default}
  translation and non-default translations were added according to
  their frequency. The version used in this paper is SVN revision
  41387.\footnote{\url{https://svn.code.sf.net/p/apertium/svn/trunk/apertium-en-es}}
\end{description}

\subsection{Performance measures}
\label{ss:measures}

This section describes the measures that will be used to evaluate the
performance of the lexical selection method proposed here: a
\emph{lexical selection performance} measure and a \emph{machine
  translation performance} measure. The main difference between the
evaluation measures is that the lexical selection error rate magnifies
the problem of lexical selection, by restricting the evaluation to
this feature. Other features of the MT system, such as the transfer rules
and morphological generation are not taken into account.

\label{sec:eval-metrics}
\subsubsection{Lexical-selection performance}
This is an intrinsic module-based evaluation of the performance of the
lexical-selection module.  It measures how well the lexical-selection
module disambiguates the output of the lexical-transfer module as
compared to a gold-standard corpus. For this task, we define a metric,
the lexical-selection error rate ({\sc ler}),
% see section~\ref{sec:ler}. 
\begin{figure}
\begin{center}
\setlength\tabcolsep{5.2pt}%
 \begin{tabular}{|l|llllll|}
   \hline
   \multicolumn{7}{|c|}{L'estiu \'{e}s una estaci\'{o} llarga}\\
   \hline
   $S$        & \emph{el} & \emph{estiu} & \emph{ser} & \emph{un} & \emph{estaci\'{o}} & \emph{llarg} \\
   \hline
   $T_s(s_i)$ & \{the\} & \{summer\} & \{be\} & \{a\} & \{station, season\} &  \{long, lengthy\} \\
   \hline 
   $T_r(s_i)$ & \{the\} & \{summer\} & \{be\} & \{a\} &  \{season\} & \{long\} \\
   \hline
   $T_t(s_i)$ & \{the\} & \{summer\} & \{be\} & \{a\} &  \{station\} & \{long\} \\
   \hline
 \end{tabular}
 \caption{An example input sentence in Catalan and the three sets of
   English translations used for calculating the lexical-selection
   error rate. The source sentence $S = (s_1, s_2, \ldots, s_{|S|})$
   has two ambiguous words, \emph{estaci\'{o}} and \emph{llarg}. There is one difference
   between the reference set $T_r(s_i)$ and the test set $T_t(s_i)$ of
   translations; as a result, the error rate for this sentence is
   50\%.}
\label{fig:ler-example}
\end{center}
\end{figure}

%\todo[inline]{FSM: LER is measured on lemmas rather than on surface
%  forms but this is not mentioned anywhere, otherwise the example in
%  Fig. \ref{fig:ler-example} is not ok.}

The lexical-selection error rate is the fraction of times the given
system chooses a translation for a word which is not found in an
annotated reference. The process uses a source-language sentence, $S =
(s_1, s_2, \ldots, s_{|S|})$ and three functions. The first function,
$T_s(s_i)$, returns all possible translations of $s_i$ according to
the bilingual dictionary. The second function, $T_t(s_i)$, returns the
translations of $s_i$ selected by the lexical-selection module:
$T_t(s_i) \subseteq T_s(s_i)$; and usually $|T_t(s_i)| = 1$. If the
lexical-selection module returns more than one translation, the first
translation is selected. \comment{FSM: How many times does this
  happens?. I would move this last sentence to an appendix.}  The
function $T_r(s_i)$ returns the set of reference translations which
are acceptable for $s_i$ in sentence $S$.\footnote{Depending on how
  the reference is built, the set returned by $T_r(s_i)$ may be
  incomplete in the sense that it may not include all possible
  acceptable translations.} For a
single sentence, we define the lexical selection error rate (LER) of
that sentence as
\begin{equation}
\mathrm{LER} = \frac{\sum_{i=1}^{|S|} \mathrm{amb}(s_i) \mathrm{diff}(T_r(s_i), T_t(s_i)}{\sum_{i=1}^{|S|} \mathrm{amb}(s_i)},
\label{eq:ler}
\end{equation}
where
\begin{equation}
\mathrm{amb}(s_i) = \left \{ \begin{matrix}
                 1 & \mbox{if } |T_s(s_i)| > 1   \\
                 0 & \mbox{otherwise} \end{matrix} \right.
\label{eq:amb}
\end{equation}
 tests if a word is ambiguous, and the  function 
\begin{equation}
\mathrm{diff}(T_r(s_i), T_t(s_i)) = \left \{ \begin{matrix}
                 1 & \mbox{if } T_r(s_i) \cap T_t(s_i) = \emptyset   \\
                 0 & \mbox{otherwise} \end{matrix} \right.
\label{eq:diff}
\end{equation}
states that there is a difference if the intersection between the set
of reference translations $T_r(s_i)$ and the set of translations from
the lexical selection module $T_t(s_i)$ is empty.  Recall that,
although $T_t(s_i)$ returns a set, this set will be a singleton, as
when the lexical-selection module returns more than one translation
the first will be selected.\footnote{In practice this does not happen
as each ambiguous word has a \emph{default} translation.}

%\todo[inline]{FSM: What happens to LER when $\sum_{i=1}^{|S|}
%  \mathrm{amb}(s_i)=0.0$?}

%\todo[inline]{FSM: LER is defined at the sentence level. It should be
%  defined at the corpus level, otherwise we would be treating short
%  sentences as the long ones, and short sentences are less likely to
%  contain ambiguous words (unfair).}

The table in Figure~\ref{fig:ler-example} gives an
overview of the inputs. In the description it is assumed that the
reference translation has been annotated by hand. However, hand
annotation is a time consuming process, and was not possible. A
description of how the reference was built is given in
section~\ref{sec:eval-corpus}.

\subsubsection{Machine translation performance} 
\comment{This part of the text needs compressing and merging} This is
an extrinsic evaluation, which ideally would test how much the system
improves as regards an approximate measurement of final translation
quality in a real system. For this task, we use the widely-used {\sc
  bleu} metric (see Section~\ref{sec:eval-metrics}). This is not ideal
for evaluating the task of a lexical selection module as the
performance of the module will depend greatly on (a) the coverage of
the bilingual dictionaries of the RBMT system in question, and (b) the
number of reference translations. It is also worth noting that
successful lexical selections may not lead to successful translations
due to inadequate transfer of morphological features. For example, if
we translate the Spanish phrase \emph{La gente le dice que no venga}
`The people tell him not to come', where \emph{decir} can be
translated as `say' or `tell' in English, and our machine translation
system generates `The people \emph{tells} him not to come', it would
be counted the same as if it were `The people \emph{says} him not to
come', although the lexical selection made was more
adequate. Generation errors --- where a word is found in the bilingual
dictionary, but not in the target language morphological dictionary
--- may also lead to the same problem.  The {\sc bleu} metric \citep{papineni02} is
included only as it is commonly used to evaluate MT systems. Other
metrics with similar performance to {\sc bleu} such as {\sc meteor}
\citep{lavie09} and {\sc nist} \citep{doddington02} are not included
as we have chosen to focus on lexical-selection performance.

%\todo[inline]{FSM: If we focus on lexical-selection performance may be
%  we should not give BLEU. Anyway, I think we can give BLEU and PER
%  (which does not take order into account) but computed using lemmas
%  instead of surface forms. }

%The {\sc bleu} (Bilingual Evaluation Understudy, \cite{papineni02})
%metric is an algorithm for evaluating the quality of text which has
%been machine-translated from one natural language to another. Quality
%is considered to be the correspondence between a machine's output and
%that of a human. The central idea is that the closer a machine
%translation is to a human translation, the better it is
%\citep{papineni02}.  BLEU was one of the first metrics that was
%reported to achieve a high correlation with human judgements of
%quality \citep{papineni02,coughlin03} and remains one of the most
%popular automated metrics.\comment{FSM: Why such a lengthy description
%  of BLEU? This description (if it is to be kept, I prefer not) should
%  appear before in this section. Same applies to the next paragraph.}
%
%Scores are calculated over a whole document --- a set of sentences ---
%by comparing them with a set of reference
%translations. Intelligibility or grammatical adequacy are not
%explicitly taken into account. BLEU is designed to approximate human
%judgement at a corpus level, and performs badly if used to evaluate
%the quality of individual sentences. The metric also does not
%correlate with human judgements when ranking systems based on
%different technologies and is recommended for tracking improvements in
%performance over different configurations of the same system
%\citep{callison06}. Additionally, \cite{denkowski12} report that it
%does not detect post-edition operations that improve translation
%quality.

\subsubsection{Confidence intervals}
Confidence intervals for both metrics will be calculated through the
\emph{bootstrap resampling} method~\citep{efron94} as described
by~\citet{koehn04} and \citet{zhang04}.\footnote{Broadly speaking,
  this involves iteratively taking a random sample (with replacements)
  from the test set and computing the score for that sample. The
  scores are sorted and, presuming we want the 95\% confidence
  interval, the top- and bottom-2.5\% are removed. The highest and
  lowest remaining are given as the interval. For more detail, refer
  to the references provided.} In all cases, bootstrap resampling will
be carried out for 1,000 iterations. Where the $p = 0.05$ confidence
intervals overlap, we will also perform paired bootstrap resampling
\citep{koehn04}.

\subsection{Corpora}
\label{sec:eval-corpus}
